Statically Typed String Sanitation Inside a Python

### 2014-08-14 sanitation-psp14/sanitation-psp14-short.tex
More editing.
Cleaned up tr rules; various other edits.
Removed counter magic from the tr rule figure
Removed contexts

Waiting for a language designer to build this feature in is unsatisfying. Moreover, we would also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language requires empirically demonstrating that it is useful, but this is difficult to do if developers have no way to use it in practice. As such, we take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.

@@

<$e$> ::= 
      $x$ \hfill source terms \alt 
      $\lambda x {:} \sigma . e$ \alt
      $e(e)$ \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{e}{e}$ \alt
      $\sisubst{r}{e}{e}$ \alt
      $\coerce{r}{e}{x.e}{e}$
\caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
\end{grammar}
\end{figure}

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash S : \sigma}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
\begin{mathpar}
\inferrule[S-T-Abs]
{\sctx, x : \sigma_1 \vdash e : \sigma_2}
{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}

\inferrule[S-T-App]
{\sctx \vdash e_1 : \sigma_1 \\ \sctx \vdash e_2 : \sigma_1 \rightarrow \sigma}
{\sctx \vdash e_1 e_2 : \sigma}

@@

\begin{figure}[t]\label{fig:lambdap}
\small
$\fbox{\inferrule{}{\tctx \vdash P : \theta}}$
~~~~$\tctx ::= \emptyset \pipe \tctx, x : \theta$

\begin{mathpar}
\inferrule[P-T-Abs]
{\tctx, x : \iota_1 \vdash e : \iota_2}
{\tctx \vdash \lambda x.e : \iota_1 \rightarrow \iota_2}

\inferrule[P-T-App]
{\tctx \vdash e_1 : \iota_1 \\ \tctx \vdash e_2 : \iota_1 \rightarrow \iota}
{\tctx \vdash e_1 e_2 : \iota}

@@

\section{Implementation in {Atlang}}\label{atlang}
In the previous section, we specified a type system and a translation semantics to a language containing only strings and regular expressions. In this section, we take Python to be such a target language. Python does not have a static type system, however, so to implement these semantics, we will leverage \verb|atlang|, an extensible type system for Python (being developed by the authors). By using \verb|atlang|, we can implement these semantics as a library, rather than as a dialect of Python.

\begin{figure}\label{fig:atexample}
\begin{lstlisting}
from atlib import fn, string_in

@@

def main(db):
  input = sanitize(user_input())
  results = db.execute(results_query(input))
  return results_div(input) + format(results)
\end{lstlisting}
\vspace{-10px}
\caption{Regular string types in atlang, a library that enables static type checking for Python. }
\end{figure}\label{fig:atimpl}
\begin{figure}
\begin{lstlisting}
class string_in(atlang.Type):
  def __init__(self, rx):
    rx = rx_normalize(rx)
    atlang.Type.__init__(idx=rx)

@@

  def check_Coerce(self, ctx, node, other_t):
    # coercions can only be defined between 
    # types with the same type constructor, 
    if rx_sublang(other_t.idx, self.idx):
      return other_t
    else: raise atlang.TypeError("...", node)
\end{lstlisting}
\vspace{-10px}
\caption{Implementation of the \texttt{string\_in} type constructor in atlang.}
\end{figure}

>>>>

Waiting for a language designer to build this feature in is unsatisfying in practice. Moreover, we also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language benefits from empirical demonstrations that it is useful, but this is difficult to do if developers have no way to use it in practice. As such, we take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.

@@

<$e$> ::= 
      $x$ \hfill source terms \alt 
      $\lambda x {:} \sigma . e$ \alt
      $e(e)$ \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{e}{e}$ \alt
      $\sisubst{r}{e}{e}$ \alt
      $\rcoerce{r}{e}$ \alt
      $\rcheck{r}{e}{x.e}{e}$
\caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
\end{grammar}
\end{figure}

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
\begin{mathpar}
\inferrule[S-T-Var]
{ x:\sigma \in \sctx }
{ \sctx \vdash s:\sigma}

\inferrule[S-T-Abs]
{\sctx, x : \sigma_1 \vdash e : \sigma_2}
{\sctx \vdash \lambda x:\sigma_1.e : \sigma_1 \rightarrow \sigma_2}

\inferrule[S-T-App]
{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
{\sctx \vdash e_1(e_2) : \sigma}

@@

\begin{figure}[t]\label{fig:lambdap}
\small
$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$

\begin{mathpar}
\inferrule[P-T-Var]
{ x:\tau \in \tctx }
{ \tctx \vdash x:\tau }

\inferrule[P-T-Abs]
{\tctx, x : \iota_1 \vdash e : \iota_2}
{\tctx \vdash \lambda x:\iota_1.e : \iota_1 \rightarrow \iota_2}

\inferrule[P-T-App]
{\tctx \vdash e_1 : \iota_2 \rightarrow \iota \\ \tctx \vdash e_2 : \iota_2}
{\tctx \vdash e_1(e_2) : \iota}

@@

\section{Implementation in {Atlang}}\label{atlang}
In the previous section, we specified a type system and a translation semantics to a language containing only strings and regular expressions. In this section, we take Python to be such a target language. Python does not have a static type system, however, so to implement these semantics, we will leverage \verb|atlang|, an extensible type system for Python (being developed by the authors). By using \verb|atlang|, which builds on Python's quotations and reflection facilities, we can implement these semantics as a library, rather than as a new dialect of Python.

\begin{figure}
\begin{lstlisting}
from atlib import fn, string_in

@@

def main(db):
  input = sanitize(user_input())
  results = db.execute(results_query(input))
  return results_div(input) + format(results)
\end{lstlisting}
\vspace{-10px}
\caption{Regular string types in atlang, a library that enables static type checking for Python. }\label{fig:atexample}
\end{figure}
\begin{figure}
\begin{lstlisting}
class string_in(atlang.Type):
  def __init__(self, rx):
    rx = rx_normalize(rx)
    atlang.Type.__init__(idx=rx)

@@

  def check_Coerce(self, ctx, node, other_t):
    # coercions can only be defined between 
    # types with the same type constructor, 
    if rx_sublang(other_t.idx, self.idx):
      return other_t
    else: raise atlang.TypeError("...", node)
\end{lstlisting}
\vspace{-10px}
\caption{Implementation of the \texttt{string\_in} type constructor in atlang.}
\label{fig:atimpl}\end{figure}
